---
title: "Assignment 6: Using Sobol with an ODE"
author: Isabella Segarra 
format:
  html:
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
  error: false
---
## Sensitivity Analysis of Forest Growth Model

```{r}
# Library set-up 
library(tidyverse)
library(deSolve)
library(here)
library(sensitivity)
set.seed(230)
```

### Forest Growth Function
1.  Implement this model in R (as a differential equation)
This is our `odefunc`.
```{r}
source(here("R/forest_growth_func.R")) 
```
Also view function below. 
```{r}
#'  Forest Growth 
#' @param time time since start
#' @param P population
#' @param parms - as list with two values, r, K
#' @param r exponential growth rate before before canopy closure
#' @param g growth rate for logistic phase
#' @param K carrying capacity 
#' @return derivative of population with time

# three parameters to include: 
# output: dC
# first function
# C: size of forest (in carbon terms)

forest_growth <- function(time, C, parms, thresh = 50) {
  #Below threshold
  if (C < thresh) {
    dC <- parms$r * C 
  # Above threshold 
  } else if (C >= thresh) {
    dC <- parms$g * (1 - C / parms$K)
  }
  return(list(dC))
}
  
```

### Ode solver 
2. Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kg/C, and using the following parameters:

-   canopy closure threshold of 50 kgC
-   K = 250 kg C (carrying capacity)
-   r = 0.01 (exponential growth rate before before canopy closure)
-   g = 2 kg/year (linear growth rate after canopy closure)
```{r}
# Starting time 300 years
tm <- seq(from = 1, to = 300) # 300 years was crashing 
# Threshold 
#thresh <- 50 
# Initial canopy closure
c_initial <- 10
# Other parameters 
parms <- list(r = 0.01, K = 250, g = 2)

# Run ODE solver 
result <- ode(y = c_initial, times = tm, func = forest_growth, parms = parms)
# Assign column names 
colnames(result) <- c("time", "dC")
result_df <- as.data.frame(result)

# Plot results
ode_300 <- result_df %>% 
ggplot(aes(time, dC)) +
  geom_line(color = "forestgreen") +
  labs(x = "Year",
       y = "Forest Size (kgC)",
       title = "Forest Growth For 300 Years",
       subtitle = "Change over time when canompy closure, carrying capacity, and growth rates\n are all held constant."
       ) +
  theme_classic() 

# View plot
ode_300

```
### Run Sobol global
3. Run a sobol global (vary all parameters at the same time) sensitivity analysis that explores how the estimated **maximum forest size** (e.g maximum of $C$ 300 years, varies with these parameters

-   pre canopy closure growth rate ($r$)
-   post-canopy closure growth rate ($g$)
-   canopy closure threshold and carrying capacity($K$)

Assume that parameters are all normally distributed with means as given above and standard deviation of 10% of mean value
```{r}
# Vary parameters

# Population size 
np <- 2000

# Parameters 
g <- rnorm(mean = 2, sd = 0.2, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
thresh <- rnorm(mean = 50, sd = 5, n = np)
K <-rnorm(mean = 250, sd = 25, n = np)

X1 <- cbind.data.frame(post_growth_rate = g, pre_growth_rate = r, thresh = thresh, carrying_capacity = K)

# repeat to get our second set of samples
g <- rnorm(mean = 2, sd = 0.2, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
thresh <- rnorm(mean = 50, sd = 5, n = np)
K <-rnorm(mean = 250, sd = 25, n = np)

X2 <- cbind.data.frame(post_growth_rate = g, pre_growth_rate = r, thresh = thresh, carrying_capacity = K)

# create our sobel object and get sets of parameters for running the model

sens_forest <- sobolSalt(model = NULL, X1, X2, nboot = 300)
# Assign column names 
colnames(sens_forest$X) <- c("g", "r", "thresh", "K")
# View sobol object 
head(sens_forest$X)
```
### Compute growth metrics function
```{r}
# Turn computing metrics into a function
compute_growth_metrics <- function(result) {
  # first add maximum population
  maxgrowth <- max(result$C)
  # return results
  return(list(maxgrowth = maxgrowth))
}
```

### Wrapper function
This function will run the ode solve, compute the growth metrics, and return results. 
```{r}
p_wrapper <- function(g, r, K, thresh, c_initial, tm, odefunc, metricfunc) {
  # create parameter list
  parms <- list(g = g, r = r,  thresh = thresh, K = K)
  
  # run ode
  result <- ode(y = c_initial, times = tm, func = odefunc, parms = parms, method="euler")
  
  # results as a data frame
  result <- data.frame(time=result[,1], C=result[,2])
  
  # compute metrics
   metrics <- metricfunc(as.data.frame(result))
   
  return(metrics)
}

# Test wrapper function with data 
p_wrapper(r = 0.01, K = 250, g = 1,thresh = 50,c_initial = c_initial, tm = tm, odefunc = forest_growth, metricfunc = compute_growth_metrics)

```
### Run wrapper function
```{r}
allresults_wrap <- as.data.frame(sens_forest$X)  %>%  pmap(p_wrapper,
                                                          c_initial = c_initial,
                                                          tm = tm,
                                                          odefunc = forest_growth,
                                                          metricfunc = compute_growth_metrics)

# extract  results from pmap into a data frame
allres <- allresults_wrap %>%   map_dfr(`[`, c("maxgrowth"))

```
5.  Graph the results of the sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (**S** and **T**).
```{r}
ggplot(allres, aes(y = maxgrowth))+ 
  geom_boxplot() + 
  theme_minimal() + 
  labs(title = "Maximum Forest Size (C)",
       y = "Max size (C)") +
  theme_classic() 
  theme(axis.text.x = element_blank()) 
```

## Determine first and second order indices. 
```{r}
# Tell the sensitivity analysis object 
sens_C_maxgrowth <- sensitivity::tell(sens_forest, allres$maxgrowth)

# First-order indices (main effect)
rownames(sens_C_maxgrowth$S) <-  c("g", "r", "thresh", "K")
sens_C_maxgrowth$S # K 


# total sensitivity index -note that this partitions the output variance - so values sum to 1
rownames(sens_C_maxgrowth$T) <-  c("g", "r", "thresh", "K")
sens_C_maxgrowth$T # 

```
### Discussion
**In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about what parameters climate change might influence)**

Based on the Sobol sensitivity analysis, maximum forest size is most sensitive to the post-canopy closure growth rate (r), followed by carrying capacity (K) in the second order. This makes sense given that forests spend the majority of 300 years in the post-canopy phase. Climate change could shift these dynamics by altering post-canopy growth rates or carrying capacity through changes in temperature, precipitation, or competition among other trees. 