---
title: "EDS-230 Assignment 7: Two variable ODE Alternative Assignment"
author: Isabella Segarra 
format:
  html:
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
  error: false
---
## Vernal Pool Fairy Shrimp Population Dynamics 
The model for this assignment is a two variable dynamic model on water depth and fairy shrimp (invertebrate) populations in vernal pools. 

```{r}
#| label: set-up

# Load packages 
library(tidyverse)
library(deSolve)
library(here)
library(ggplot2)
library(sensitivity)
# Set seed for reproducibility
set.seed(230)

```

## Water-depth and invert pop function
- water depth increases with precipitation, decreases with evaporation
- invertebrate density grows logistically, carrying capacity is correlated with water depth, populations fall once water depth reaches a minimum level (mortality)
- when pool water drops below the threshold (W < W_thresh) then this value evaluates and mortality switches on; above the threshold it evaluates to 0 and mortality is not applied 
```{r}
#' Vernal Pool model 
#'
#' function computes the rate of change of populations for invertebrate populations with water depth of vernal pools as a threshold. 
#' @param t  time (days)
#' @param state numeric vector 
#' @param params list of model parameters (p = precipitation input (mm/day), e = evaporation rate, k = carrying capacity, r = intrisic growth rate of invertebrates, m = dessication mortality rate, w_thresh = water depth trehshold (mm))
#' @return A list containing the following components:
#'  \describe{
#' \item{dW}{rate of change of water depth}
#' \item{dN}{rate of change of populaiton}
#' }

vernal_pool_func <- function(t, state, params){
  W <- state["W"]
  N <- state["N"]
  #Water depth: constant precip input, evaporation and infiltration losses
  with(as.list(params), {
    dW <- P - e * W

    # Carrying capacity correlates to W (water) level
    k_w <- k * W
    
    #Invertebrate density
   dN <- r * N * (1 - N / k_w) - m * N * (W < W_thresh) 
    list(c(dW, dN))
  }) 
}

```

## Run ODE 
```{r}
# Define parameters

# time (180 days ~ wet season)
tm <- seq(from = 0, to = 180)
# Initial water and population 
y0 <- c(W = 50, N = 5)

# Params
params <- list(
  P = 1, # precipitation, originally 0.5 --> population crashed at day 50 
  e = 0.05, # evaporation
  k = 10, # carrying capacity
  r = 0.1, #growth rate
  m = 0.5, # dessication mortality rate
  W_thresh = 20 # depth threshold 
)

result <- ode(y = y0, times = tm, func = vernal_pool_func, parms = params)

result_df <- as.data.frame(result)


result_df %>%
  ggplot(aes(W, N)) +
  geom_path() +
  labs(x = "Water Depth (mm)",
       y = "Fairy Shrimp Population",
       title = "Fairy shrimp populations with water depth (mm)") +
  theme_classic()

```
## Sensitivity analaysis - Vary parameters

### Sobol 
Vary all parameters at the same time. Determine which parameters are the most sensitive in this model. 
```{r}
# Define population size 
np <- 2000

# First sample
# why not rnorm? gives negative values that Sobol does not like 
P     <- rnorm(mean = 2,  sd = 0.05,  n = np)
e     <- rnorm(mean = 0.05, sd = 0.005, n = np)
k     <- rnorm(mean = 10,   sd = 1,     n = np)
r     <- rnorm(mean = 0.1,  sd = 0.01,  n = np)
m     <- rnorm(mean = 0.5,  sd = 0.05,  n = np)
W_thresh <- rnorm(mean = 20, sd = 2,    n = np)
X1 <- cbind.data.frame(P, e, k, r, m, W_thresh)

# Re-run for second sample
P     <- rnorm(mean = 0.5,  sd = 0.05,  n = np)
e     <- rnorm(mean = 0.05, sd = 0.005, n = np)
k     <- rnorm(mean = 10,   sd = 1,     n = np)
r     <- rnorm(mean = 0.1,  sd = 0.01,  n = np)
m     <- rnorm(mean = 0.5,  sd = 0.05,  n = np)
W_thresh <- rnorm(mean = 20, sd = 2,    n = np)
X2 <- cbind.data.frame(P, e, k, r, m, W_thresh)

# Run sobol 
sens_pool <- sobolSalt(model = NULL, X1, X2, nboot = 300)
colnames(sens_pool$X) <- c("P", "e", "k", "r", "m", "W_thresh")

head(sens_pool$X)

# Wrapper function for iterating ODE for all params

p_wrapper <- function(P, e, k, r, m, W_thresh, y0, times, odefunc) {
  # create parameter list
  params <- list(P = P, e = e, k = k, r = r, m = m, W_thresh = W_thresh)
  
  # run ode
  result <- ode(y = y0, times = tm, func = vernal_pool_func, parms = params, method = "euler")
  
  # results as a data frame
  result_df <- data.frame(result)
  
  # Return population at day 180 
  return(result_df$N[nrow(result)])
}

# Test p-wrapper
p_wrapper(P = 2, e = 0.2, k = 20, r = 0.2, m = 0.2, W_thresh = 10, y0 = y0, times = tm)

# Run wrapper for all 
allresults <- as.data.frame(sens_pool$X) %>%
  pmap(p_wrapper, y0 = y0, times = tm, odefunc = vernal_pool_func)

allres <- data.frame(final_N = unlist(allresults))

# Tell the sensitivity analysis object
sens_pool <- sensitivity::tell(sens_pool, allres$final_N)

# First-order indices
rownames(sens_pool$S) <- c("P", "e", "k", "r", "m", "W_thresh")
sens_pool$S

# Total sensitivity index
rownames(sens_pool$T) <- c("P", "e", "k", "r", "m", "W_thresh")
sens_pool$T


```
As expected, the most sensitive parameter is precipitation. This parameter controls most of the variation in the model. 


## Stability 
For this stability analysis, I want to find the minimum precipitation of fairy shrimp a pool needs to have a sustaining population for the next year. 
```{r}
# Chain ODEs per year together 
interannual_ode <- function(n_years, y0, params, times) {
  y_current <- y0
  yearly_peaks <- vector("list", n_years)
  
  for (yr in 1:n_years) {
    result <- ode(y = y_current, times = times, func = vernal_pool_func,
                  parms = params, method = "euler")
    result_df <- as.data.frame(result)
    peak_N <- max(result_df$N)
    y_current <- c(W = unname(y0["W"]), N = result_df$N[nrow(result_df)])
    yearly_peaks[[yr]] <- data.frame(year = yr, peak_N = peak_N)
  }
  bind_rows(yearly_peaks)
}

# Minimum precipitation needed for stable population
P_seq <- seq(0.1, 3, by = 0.1)

Pop_stability <- map_dfr(P_seq, function(p_val) {
  p_new <- params
  p_new$P <- p_val
  
  peaks <- interannual_ode(n_years = 10, y0 = y0, params = p_new, times = tm)
  
  data.frame(P = p_val, avg_peak = mean(peaks$peak_N))
})

Pop_stability %>%
  ggplot(aes(P, avg_peak)) +
  geom_point() +
  labs(x = "Precipitation (mm/day)", y = "Average Annual Peak Population",
       color = "Stable", title = "Minimum Precipitation for Fairy Shrimp Population Stability") +
  theme_classic()
```
Precipitation below 1 mm/day leads to close to zero average annual populations for fairy shrimp. Once precipitation rises above 1 mm/day, then the populations increase linearly. From an ecological standpoint, this means that fairy shrimp are highly sensitive to daily precipitation. From a conservation standpoint, fairy shrimp populations could be severely impacted by drought, however they have been shown to stay dormant within the soil for long periods of aridity. 


